---
title: "Exporting data for Cellucid"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Exporting data for Cellucid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The `cellucid` package exports single-cell data (embeddings, metadata, gene
expression, and connectivity) to the binary/JSON format consumed by the
Cellucid WebGL viewer.

The main entry point is `cellucid_prepare()` (also exported as `prepare()` to
mirror the Python API).

## Minimal example

```{r}
library(cellucid)

latent <- matrix(c(0, 0,
                   1, 1,
                   2, 2),
                 ncol = 2, byrow = TRUE)

obs <- data.frame(
  cluster = factor(c("A", "A", "B")),
  score = c(0.1, 0.2, 0.3)
)

umap2 <- matrix(c(0, 0,
                  1, 0,
                  0, 1),
                ncol = 2, byrow = TRUE)

expr <- matrix(c(0, 1,
                 2, 3,
                 4, 5),
               nrow = 3, ncol = 2, byrow = TRUE)

var <- data.frame(symbol = c("G1", "G2"))
rownames(var) <- var$symbol

out_dir <- file.path(tempdir(), "cellucid_vignette_export")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

cellucid_prepare(
  latent_space = latent,
  obs = obs,
  var = var,
  gene_expression = expr,
  X_umap_2d = umap2,
  out_dir = out_dir,
  centroid_min_points = 1,
  force = TRUE
)

list.files(out_dir, recursive = TRUE)
```

## Using Seurat

Seurat objects typically store expression as **genes × cells**, but
`cellucid_prepare()` expects **cells × genes**, so transpose before passing.

```{r, eval=FALSE}
if (requireNamespace("Seurat", quietly = TRUE) &&
    requireNamespace("Matrix", quietly = TRUE)) {

  seu <- your_seurat_object

  umap2 <- Seurat::Embeddings(seu, "umap")[, 1:2, drop = FALSE]
  latent <- tryCatch(
    Seurat::Embeddings(seu, "pca"),
    error = function(e) umap2
  )

  obs <- seu@meta.data

  expr_genes_by_cells <- Seurat::GetAssayData(seu, slot = "data")
  expr_cells_by_genes <- Matrix::t(expr_genes_by_cells)

  var <- data.frame(symbol = rownames(expr_genes_by_cells))
  rownames(var) <- var$symbol

  cellucid_prepare(
    latent_space = latent,
    obs = obs,
    var = var,
    gene_expression = expr_cells_by_genes,
    X_umap_2d = umap2,
    out_dir = "exports/seurat_export",
    force = TRUE
  )
}
```

## Using SingleCellExperiment

SingleCellExperiment assays are also typically **genes × cells**, so transpose
before passing.

```{r, eval=FALSE}
if (requireNamespace("SingleCellExperiment", quietly = TRUE) &&
    requireNamespace("SummarizedExperiment", quietly = TRUE) &&
    requireNamespace("Matrix", quietly = TRUE)) {

  sce <- your_sce_object

  umap2 <- SingleCellExperiment::reducedDim(sce, "UMAP")
  latent <- tryCatch(
    SingleCellExperiment::reducedDim(sce, "PCA"),
    error = function(e) umap2
  )

  obs <- as.data.frame(SummarizedExperiment::colData(sce))

  expr_genes_by_cells <- SummarizedExperiment::assay(sce, "logcounts")
  expr_cells_by_genes <- Matrix::t(expr_genes_by_cells)

  var <- as.data.frame(SummarizedExperiment::rowData(sce))
  rownames(var) <- rownames(sce)

  cellucid_prepare(
    latent_space = latent,
    obs = obs,
    var = var,
    gene_expression = expr_cells_by_genes,
    X_umap_2d = umap2,
    out_dir = "exports/sce_export",
    force = TRUE
  )
}
```

## Output format (high level)

The export directory contains:

- Embeddings: `points_1d.bin`, `points_2d.bin`, `points_3d.bin` (float32)
- Cell metadata: `obs_manifest.json` + `obs/` binaries
- Gene expression: `var_manifest.json` + `var/` binaries (one file per gene)
- Connectivity (optional): `connectivity_manifest.json` + `connectivity/` binaries
- Dataset metadata: `dataset_identity.json`
