---
title: "Installing Cellucid for R"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Installing Cellucid for R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Current version and availability

**The active Cellucid for R source and documentation version is 0.9.1.** It is
the CRAN submission release. The
[CRAN package index](https://cran.r-project.org/web/packages/available_packages_by_name.html)
is authoritative for registry availability. Use the CRAN command below when
the index lists `cellucid` 0.9.1; otherwise use the GitHub command.

## Requirements

- R 4.3.0 or newer
- An internet connection while installing the package
- The `remotes` package only when installation from GitHub is needed

Cellucid is implemented in R and does not compile native code. The optional
`Matrix` package is recommended when exporting sparse expression matrices or
connectivity graphs.

Check your R version with:

```{r, eval=FALSE}
R.version.string
```

## Install version 0.9.1 from CRAN

When the CRAN package index lists `cellucid` 0.9.1, install it with:

```{r, eval=FALSE}
install.packages("cellucid")
```

If the index does not list version 0.9.1 yet, install the active source from
the official GitHub repository:

```{r, eval=FALSE}
install.packages("remotes")
remotes::install_github("theislab/cellucid-r")
```

Install the recommended sparse-matrix support when your data uses sparse
expression or connectivity matrices:

```{r, eval=FALSE}
install.packages("Matrix")
```

Restart R after upgrading an installation that was already loaded in the
current session.

## Update an existing installation

Install again from the source currently available to you. Use CRAN when its
package index lists version 0.9.1:

```{r, eval=FALSE}
install.packages("cellucid")
```

Otherwise update directly from the official source repository:

```{r, eval=FALSE}
remotes::install_github("theislab/cellucid-r")
```

Restart R after the update so the current session does not retain the package
version that was loaded earlier.

## Verify the installation

Load Cellucid and confirm that R reports the active version:

```{r, eval=FALSE}
library(cellucid)
packageVersion("cellucid")
#> [1] '0.9.1'
```

You can then open the preparation function reference:

```{r, eval=FALSE}
?cellucid_prepare
```

Continue with [Exporting data for Cellucid](cellucid.html) for a complete
example, or use the
[R package guide](https://cellucid.readthedocs.io/en/latest/user_guide/r_package/index.html)
for Seurat and SingleCellExperiment recipes.

## Next step

Once the package is installed, prepare an export with `cellucid_prepare()` and
open the resulting directory in the
[Cellucid web app](https://www.cellucid.com). The
[function reference](../reference/cellucid_prepare.html) documents every input
and validation rule.
